Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market

### 2023-07-02 paper/reviewer_response.tex

The author states more than once that the first infections were in November or earlier.   The statement in the introduction was, “A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is at least a month after the first human infections in Wuhan.”  While this is probably correct, it should not be stated as a fact.  I looked up most of the reference cited and none of them were so definitive.  Among the documents was an unclassified government summary which concluded that the virus ‘probably’ emerged no later than November, 2019.  Without the ‘probably’ or some other modifier, what is written in the manuscript is misrepresenting the government document.

\response{
I have added the qualifier ``probably'' to the references in both the Introduction and Discussion about how the first cases probably occurred no later than November of 2019.
I have also added several more citations to papers that use various methods to try to date the first human infections.
All of these papers make a point estimate of the date of the first human infection that is no later than November of 2019, although some of them do estimate credible or confidence intervals that extend into early December of 2019.
}

Conspicuously missing from the introduction is any discussion of the epidemiology papers that suggest that the virus originated in the market.  A description could come with commentary, but to simply ignore the work makes the manuscript come across as biased.

\response{
This is a good suggestion.
I have added new text (third paragraph of revised Introduction) that discusses the interpretations of epidemiological data about the origin of the virus.
}

@@

I don’t love the 20\% cutoff.  I understand why the table in the paper needs to have the 20\% cutoff to make it a manageable size, but the complete dataset is included as supplementary, so there isn’t any reason for the text to focus on the arbitrary 20\% as much as it does.  I suppose it isn’t essential to do away with it, but people who don’t like the paper are going to say that this is cherry picking.

The abstract states ``For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.''  After reading the manuscript this sentence makes sense, but a lot of people are probably only going to read the abstract, so the 20\% cutoff seems would seem arbitrary, and the 1 in 200M has not context.  In the abstract makes it sounds ridiculously low, but many of the samples had viral copies in the single digits.  If this sentence is included I think it should be reworded or put into context (elaborating on what a `typical' number of reads is).

\response{
As the reviewer notes, the paper contains in the figures and a supplementary table analysis of the association between SARS-CoV-2 content and mitochondrial material composition for all samples without any cutoff, and the 20\% cutoff was applied just to make Table 1 fit in the main text as the full table is too large.

However, I see the reviewer's point that the 20\% cutoff is reasonable for creating the table, but that the main text should not rely heavily on that particular cutoff in describing the results as otherwise the reader may get the incorrect impression that the 20\% cutoff is somehow integral to the analysis.
I also agree that the abstract should clearly emphasize how nearly all samples contain low SARS-CoV-2 content.
I have made revisions to the text to reflect these suggestions. 
The abstract now emphasizes the low overall SARS-CoV-2 content and relatively de-emphasizes the 20\% cutoff with the new wording below.
Similar changes in wording have been made to the discussion text.
\begin{quote}
\textit{
However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
}
\end{quote}
}

Related to the above point, it seems to me that the manuscript missed an opportunity to drive home the point that the data cannot be used to infer viral source.  The one way that one could get a smoking gun from this kind of data is if there was a sample that had a whole lot of viral material, and basically only one species associated with it (which didn’t happen).  In fact, the two samples with the highest SC2 levels didn’t have high levels of raccoon dog, human or any other susceptible species.  Using that as a foil one could point out that the positive with the most raccoon dog had only once SC2 copy, while the samples with the most SC2 copies ($>$1000 times more) had no prevalent candidate hosts associated with it at all.  Just a thought.

### 2023-07-02 paper/paper.tex
Reviewer point 3: de-emphasize 20% cutoff

\begin{abstract}
The role of the Huanan Seafood Market in the early SARS-CoV-2 outbreak remains unclear.
Recently the Chinese CDC released data from deep sequencing of environmental samples collected from the market after it was closed on January-1-2020~\citep{liu2023surveillance}.
Prior to this release, \citet{crits2023genetic} analyzed data from a subset of the samples.
Both studies concurred that the samples contained genetic material from a variety of species, including some like raccoon dogs that are susceptible to SARS-CoV-2.
However, neither study systematically analyzed the relationship between the amount of genetic material from SARS-CoV-2 and different animal species.
Here I implement a fully reproducible computational pipeline that jointly analyzes the number of reads mapping to SARS-CoV-2 and the mitochondrial genomes of chordate species across the full set of samples.
I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species.
For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
These results suggest that while metagenomic analysis of the environmental samples is useful for identifying animals or animal products sold at the market, co-mingling of animal and viral genetic material is unlikely to reliably indicate whether any animals were infected by SARS-CoV-2.
\end{abstract}

@@

Third, I perform the first comprehensive analysis of the association between the abundance of SARS-CoV-2 and mitochondrial genetic material across all environmental samples.
This analysis reveals that the greatest co-mingling of viral and animal material involves species that were almost certainly not infected by SARS-CoV-2, such as fish (e.g., largemouth bass, catfish) and livestock (e.g., cows, sheep, goats).
Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: 13 of the 14 samples with at least 20\% of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Likewise, none of the six samples with at least 20\% bamboo rat mitochondrial material contain any SARS-CoV-2 reads.
These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.
However, they are somewhat inconsistent with related media articles that emphasized co-mingling of raccoon dog and viral material~\citep{wu2023atlantic,mueller2023nytimes}---in fact, raccoon dogs are one of the species with the least co-mingling of their genetic material and SARS-CoV-2.
The basic finding that SARS-CoV-2 material is not associated with material from susceptible non-human species sold live at the market is largely robust to examining subsets of the samples (such as just those collected on specific dates), and I provide interactive plots to facilitate visualizing the data in different ways.

>>>>

\begin{abstract}
The role of the Huanan Seafood Market in the early SARS-CoV-2 outbreak remains unclear.
Recently the Chinese CDC released data from deep sequencing of environmental samples collected from the market after it was closed on January-1-2020~\citep{liu2023surveillance}.
Prior to this release, \citet{crits2023genetic} analyzed data from a subset of the samples.
Both studies concurred that the samples contained genetic material from a variety of species, including some like raccoon dogs that are susceptible to SARS-CoV-2.
However, neither study systematically analyzed the relationship between the amount of genetic material from SARS-CoV-2 and different animal species.
Here I implement a fully reproducible computational pipeline that jointly analyzes the number of reads mapping to SARS-CoV-2 and the mitochondrial genomes of chordate species across the full set of samples.
I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
These results suggest that while metagenomic analysis of the environmental samples is useful for identifying animals or animal products sold at the market, co-mingling of animal and viral genetic material is unlikely to reliably indicate whether any animals were infected by SARS-CoV-2.
\end{abstract}

@@

Third, I perform the first comprehensive analysis of the association between the abundance of SARS-CoV-2 and mitochondrial genetic material across all environmental samples.
This analysis reveals that the greatest co-mingling of viral and animal material involves species that were almost certainly not infected by SARS-CoV-2, such as fish (e.g., largemouth bass, catfish) and livestock (e.g., cows, sheep, goats).
Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: for instance, 13 of the 14 samples with at least a fifth of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.
However, they are somewhat inconsistent with related media articles that emphasized co-mingling of raccoon dog and viral material~\citep{wu2023atlantic,mueller2023nytimes}---in fact, raccoon dogs are one of the species with the least co-mingling of their genetic material and SARS-CoV-2.
The basic finding that SARS-CoV-2 material is not associated with material from susceptible non-human species sold live at the market is largely robust to examining subsets of the samples (such as just those collected on specific dates), and I provide interactive plots to facilitate visualizing the data in different ways.
